\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-04-25 mSystems/mSys_MandM.tex
In mSystems move section from MandM to Results (v0.0.6)
On branch master
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_results.tex

\subsubsection*{Sample selection}
We have chosen studies about relevant pathologies containing metagenomic sequencing time data series of bacterial populations from humans in different healthy and non-healthy states. We have selected only those individuals who had three or more time points of data available in databases. Metadata of each study is provided in Supplementary Tables \ref{tab:diet} to \ref{tab:LEA}. All used 16S rRNA gene sequencing except for the study of the discordant kwashiorkor twins\cite{kwashiorkor} (see Supplementary Tables \ref{tab:DH} and \ref{tab:DK}) where shotgun metagenomic sequencing (SMS) and 16S rRNA were used. In the latter case we selected to work with SMS data to show that our method is valid regardless of the source of taxonomic information. Each one of the datasets was treated as follows:

@@

\subsubsection*{Taxa level selection}
We selected genus as taxonomic level for the subsequent steps of our work. In order to ensure that, between adjacent taxonomic levels, there were not crucial differences which could still be of relevance after standardization (see Section \ref{sec:stan}), we tested two different data sets. In the former, the antibiotics study\cite{antibiotic} with 16S data, we tested the differences between genus and family levels. The latter dataset tested was the kwashiorkor discordant twins study\cite{kwashiorkor} for both genus and species taxonomic levels. The Supplementary Figures \ref{fig:taxlev1} (overview) and \ref{fig:taxlev2} (detail) plot the comparison between studies (and so, 16S and SMS) and between adjacent taxonomic levels.

@@

\subsection*{\CC\ output}
In the previous sections, we have discussed details about the methods used in \CC. In this section we review the output from the package, which aims at summarizing the as yet undescribed functionality.

\subsubsection*{Fit Plots}
Both for the unweighted fit (detailed in the section \ref{sec:unw}) and for the X-weighted fit (detailed in the section \ref{sec:X-w}), two plots are generated: the former with logarithmic scale, the latter with lineal scale. Figure \ref{fig:unwFit} shows an example of the unweighted fit plots and Figure \ref{fig:X-wFit} does the same with the X-weighted fit plots. Additionally, for the unweighted fit, a complete residues analysis is performed, and a 4-in-1 figure is generated as shown in Figure \ref{fig:unwRes}, corresponding to the fit of Figure \ref{fig:unwFit}. Among other tests, it allows to check for normality and homoscedasticity of the residues.

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_LLR_LOG.pdf}
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_LLR_LIN.pdf}
	\caption{Example of unweighted fit, shown both in logarithmic and lineal scale plots}
	\label{fig:unwFit}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_xWboot_LOG.pdf}
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_xWboot_LIN.pdf}
	\caption{The X-weighted fit log and lineal plots corresponding to the fit of Figure \ref{fig:unwFit}}
	\label{fig:X-wFit}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_LLR_RES.pdf}
	\caption{Residues analysis plot corresponding to the fit of Figure \ref{fig:unwFit}. The top-left subplot is a simple residues plot. The top-right subplot is a Normal quantiles plot with linear fitting (value of coefficient of determination is provided). The bottom-left subplot shows an accumulated ESS (Explained Sum of Squares) plot. Finally, the bottom-right subplot is a residues Normal histogram plot. This set of subplots allows to check for normality and homoscedasticity of the residues.}
	\label{fig:unwRes}
\end{figure}

\subsubsection*{Histogram Plots} 
\CC\ generates three different histogram plots:
\begin{description}
	\item[$-$Absolute frequencies plot]: This plot is useful to visually assess the validity of the time points in terms of the accumulated absolute frequency of the elements (taxa), since absolute frequencies far (much higher or much lower) from those typically observed could mean a sampling problem. As an example, Figure \ref{fig:histAFP} shows this histogram for the pre-treatment data (first 7 times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$2D deviation plot]: The 2D semi-logarithmic histogram representing deviations from the mean versus the mean itself, is a useful tool in the analysis of the stability of ranking processes in complex systems\cite{ranking}. Figure \ref{fig:hist2D} shows this plot for the data used in the fit shown in Figure \ref{fig:unwFit}.
	\item[$-$Zero relative frequency plot]: We could define the ZRF (Zero Relative Frequency, thereby ranging from 0 to 1) of an element (taxon) as the portion of times where it is zero, i.e., it is not found. Attending to all the elements (taxa), we can plot the ZRF histogram, which then lies on the horizontal axis of the plot. The vertical axis shows the number of elements (taxa), so the height of a bar represents the number of elements that have determinate ZRF. In this respect, the bar over $0.0$ counts the number of elements (taxa) that are present at every time point of the data set (aka ``core''), while the bar over $1.0$ would count the number of elements (taxa) that are never found (this bar never appears because all these ``null'' elements are automatically filtered by the code). Figure \ref{fig:histZRF} shows an example of this plot. There, we can see that $12$ taxa are present at all the time points of the time series while $9$ taxa basically appear only once. So, this plot is clearly useful to notice how the ``core'' is distributed.
\end{description}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/hist/antibiotics_h_patD_pre_AbsFreqPlot}
	\caption{Histogram with the absolute frequencies of the pre-treatment data (7 first times) of patient ``D'' in the antibiotics study\cite{antibiotic}}
	\label{fig:histAFP}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/hist/IBS_h_A_amplicons_family_hist2D.pdf}
	\caption{2D histogram deviation plot of the data used in the fit shown in Figure \ref{fig:unwFit}}
	\label{fig:hist2D}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/hist/IBS_h_A_amplicons_family_ZRFhist.pdf}
	\caption{Histogram with the relative frequency of zero for the elements (taxa) in the data used in the fit shown in Figure \ref{fig:unwFit}}
	\label{fig:histZRF}
\end{figure}

\subsubsection*{Correlation and Rank Plots} 
\CC\ generates three different plots falling under this category, as well as Excel files with the resulting matrices:
\begin{description}
	\item[$-$Elements correlation matrix]: This plot shows a correlation matrix among the elements (taxa), calculated with the time as independent variable. For these calculations, the data set is not normalized to avoid entering an additional constraint. As an example, Figure \ref{fig:corrElm} shows this matrix for the ``core'' elements (taxa) present in the pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$Times correlation matrix]: This plot presents a correlation matrix among the time points of the data set, calculated with the elements (taxa) as independent variable. Again, the data set is not normalized. Figure \ref{fig:corrTim} shows this matrix for the ``core'' elements (taxa) present in pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$Rank dynamics and stability plot]: This plot shows the variation in the rank with time for the most dominant elements (taxa) and their calculated RSI, as discussed in Section \ref{sec:RSI}. Figure \ref{fig:corrank} shows this plot for the elements (taxa) in the data used in the fit shown in Figure \ref{fig:unwFit}.
\end{description}

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{results/corrank/antibiotics_h_patD_pre_CORE_ElementsCorr}
	\caption{Element correlation plot of the pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic} for ``core'' taxa}
	\label{fig:corrElm}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{results/corrank/antibiotics_h_patD_pre_CORE_TimesCorr}
	\caption{Times correlation plot of the pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic} for ``core'' taxa}
	\label{fig:corrTim}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{results/corrank/IBS_h_A_amplicons_family_Rank}
	\caption{Matrix showing the rank variation throughout time for the most dominant elements (taxa) and their calculated RSI (as discussed in Section \ref{sec:RSI}) in the data used in the fit shown in Figure \ref{fig:unwFit}}
	\label{fig:corrank}
\end{figure}

>>>>

\subsubsection*{Sample selection}
We have chosen studies about relevant pathologies containing metagenomic sequencing time data series of bacterial populations from humans in different healthy and non-healthy states. We have selected only those individuals who had three or more time points of data available in databases. Metadata of each study is provided in Tables \ref{tab:diet} to \ref{tab:LEA}. All used 16S rRNA gene sequencing except for the study of the discordant kwashiorkor twins\cite{kwashiorkor} (see Tables \ref{tab:DH} and \ref{tab:DK}) where shotgun metagenomic sequencing (SMS) and 16S rRNA were used. In the latter case we selected to work with SMS data to show that our method is valid regardless of the source of taxonomic information. Each one of the datasets was treated as follows:

@@

\subsubsection*{Taxa level selection}
We selected genus as taxonomic level for the subsequent steps of our work. In order to ensure that, between adjacent taxonomic levels, there were not crucial differences which could still be of relevance after standardization (see Section \ref{sec:stan}), we tested two different data sets. In the former, the antibiotics study\cite{antibiotic} with 16S data, we tested the differences between genus and family levels. The latter dataset tested was the kwashiorkor discordant twins study\cite{kwashiorkor} for both genus and species taxonomic levels. The Figures \ref{fig:taxlev1} (overview) and \ref{fig:taxlev2} (detail) plot the comparison between studies (and so, 16S and SMS) and between adjacent taxonomic levels.

@@

### 2016-04-25 mSystems/mSys_results.tex
In mSystems move section from MandM to Results (v0.0.6)
On branch master
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_results.tex

@@

>>>>

%%%
\subsection*{Global results}

@@

%%%
\subsection*{DRAFT: Specific results}

\subsubsection*{Fit Plots}
Both for the unweighted fit (see Material and Methods section for details) and for the X-weighted fit (detailed in Material and Methods subsection), two plots are generated: the former with logarithmic scale, the latter with lineal scale. Figure \ref{fig:unwFit} shows an example of the unweighted fit plots and Figure \ref{fig:X-wFit} does the same with the X-weighted fit plots. Additionally, for the unweighted fit, a complete residues analysis is performed, and a 4-in-1 figure is generated as shown in Figure \ref{fig:unwRes}, corresponding to the fit of Figure \ref{fig:unwFit}. Among other tests, it allows to check for normality and homoscedasticity of the residues.

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_LLR_LOG.pdf}
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_LLR_LIN.pdf}
	\caption{Example of unweighted fit, shown both in logarithmic and lineal scale plots}
	\label{fig:unwFit}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_xWboot_LOG.pdf}
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_xWboot_LIN.pdf}
	\caption{The X-weighted fit log and lineal plots corresponding to the fit of Figure \ref{fig:unwFit}}
	\label{fig:X-wFit}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_LLR_RES.pdf}
	\caption{Residues analysis plot corresponding to the fit of Figure \ref{fig:unwFit}. The top-left subplot is a simple residues plot. The top-right subplot is a Normal quantiles plot with linear fitting (value of coefficient of determination is provided). The bottom-left subplot shows an accumulated ESS (Explained Sum of Squares) plot. Finally, the bottom-right subplot is a residues Normal histogram plot. This set of subplots allows to check for normality and homoscedasticity of the residues.}
	\label{fig:unwRes}
\end{figure}

\subsubsection*{Histogram Plots} 
\CC\ generates three different histogram plots:
\begin{description}
	\item[$-$Absolute frequencies plot]: This plot is useful to visually assess the validity of the time points in terms of the accumulated absolute frequency of the elements (taxa), since absolute frequencies far (much higher or much lower) from those typically observed could mean a sampling problem. As an example, Figure \ref{fig:histAFP} shows this histogram for the pre-treatment data (first 7 times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$2D deviation plot]: The 2D semi-logarithmic histogram representing deviations from the mean versus the mean itself, is a useful tool in the analysis of the stability of ranking processes in complex systems\cite{ranking}. Figure \ref{fig:hist2D} shows this plot for the data used in the fit shown in Figure \ref{fig:unwFit}.
	\item[$-$Zero relative frequency plot]: We could define the ZRF (Zero Relative Frequency, thereby ranging from 0 to 1) of an element (taxon) as the portion of times where it is zero, i.e., it is not found. Attending to all the elements (taxa), we can plot the ZRF histogram, which then lies on the horizontal axis of the plot. The vertical axis shows the number of elements (taxa), so the height of a bar represents the number of elements that have determinate ZRF. In this respect, the bar over $0.0$ counts the number of elements (taxa) that are present at every time point of the data set (aka ``core''), while the bar over $1.0$ would count the number of elements (taxa) that are never found (this bar never appears because all these ``null'' elements are automatically filtered by the code). Figure \ref{fig:histZRF} shows an example of this plot. There, we can see that $12$ taxa are present at all the time points of the time series while $9$ taxa basically appear only once. So, this plot is clearly useful to notice how the ``core'' is distributed.
\end{description}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/hist/antibiotics_h_patD_pre_AbsFreqPlot}
	\caption{Histogram with the absolute frequencies of the pre-treatment data (7 first times) of patient ``D'' in the antibiotics study\cite{antibiotic}}
	\label{fig:histAFP}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/hist/IBS_h_A_amplicons_family_hist2D.pdf}
	\caption{2D histogram deviation plot of the data used in the fit shown in Figure \ref{fig:unwFit}}
	\label{fig:hist2D}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/hist/IBS_h_A_amplicons_family_ZRFhist.pdf}
	\caption{Histogram with the relative frequency of zero for the elements (taxa) in the data used in the fit shown in Figure \ref{fig:unwFit}}
	\label{fig:histZRF}
\end{figure}

\subsubsection*{Correlation and Rank Plots} 
\CC\ generates three different plots falling under this category, as well as Excel files with the resulting matrices:
\begin{description}
	\item[$-$Elements correlation matrix]: This plot shows a correlation matrix among the elements (taxa), calculated with the time as independent variable. For these calculations, the data set is not normalized to avoid entering an additional constraint. As an example, Figure \ref{fig:corrElm} shows this matrix for the ``core'' elements (taxa) present in the pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$Times correlation matrix]: This plot presents a correlation matrix among the time points of the data set, calculated with the elements (taxa) as independent variable. Again, the data set is not normalized. Figure \ref{fig:corrTim} shows this matrix for the ``core'' elements (taxa) present in pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$Rank dynamics and stability plot]: This plot shows the variation in the rank with time for the most dominant elements (taxa) and their calculated RSI, as discussed in Section \ref{sec:RSI}. Figure \ref{fig:corrank} shows this plot for the elements (taxa) in the data used in the fit shown in Figure \ref{fig:unwFit}.
\end{description}

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{results/corrank/antibiotics_h_patD_pre_CORE_ElementsCorr}
	\caption{Element correlation plot of the pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic} for ``core'' taxa}
	\label{fig:corrElm}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{results/corrank/antibiotics_h_patD_pre_CORE_TimesCorr}
	\caption{Times correlation plot of the pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic} for ``core'' taxa}
	\label{fig:corrTim}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{results/corrank/IBS_h_A_amplicons_family_Rank}
	\caption{Matrix showing the rank variation throughout time for the most dominant elements (taxa) and their calculated RSI (as discussed in Section \ref{sec:RSI}) in the data used in the fit shown in Figure \ref{fig:unwFit}}
	\label{fig:corrank}
\end{figure}

%%%
\subsection*{Time dependence of model parameters}
